---
title: "grad_assignment"
author: "Pablo Suarez"
date: "2023-11-24"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## INTRODUCTION

Welcome to Pablo Suarez's markdown file for the JOUR772 Graduate Student
Assignment. Below, I am analyzing the Reported Sewer Overflows data set
provided by the Maryland Department of the Environment. This data set
contains information regarding reported sewer overflows across the state
of Maryland from January 1, 2005 through January 31, 2023.

An important note from the data set creators: "Although MDE requires
that all public sewer system owners or operators report overflows to us,
there may be incidents that were not reported. Note that overflow
amounts provided by the person reporting the overflow may be estimated
using best professional judgment or they may be actual readings from
flow measurement devices when available. Penalty information started
during 2013."

Data Profile, Scope and Limitations:

Rows -- 27,479

Columns -- 22

Who created and maintains the data? -- Maryland Department of the
Environment

Dataset Owner -- Andrew Gosden

What state program or law is connected to it? -- Unclear

What does each row represent? -- One overflow event

Columns: overflow type, municipality/facility (it occurred in), NPDES \#
(National Pollutant Discharge Elimination System identification number
for the facility or system), date discovered, time discovered, duration
(days, hours, minutes), location, zip code, collection-system, quantity
in gallons, net in gallons, cause, watershed, receiving waters, county,
notes, penalty collected, penalty comments, latitude, longitude

What is the chronological and geographic scope of this data? -- Reported
sewer overflows across the state of Maryland from January 1, 2005,
through January 31, 2023.

Newsworthy questions that I'd like to answer with this data set: Which
counties reported the most sewer overflows over the dataset's observed
period of time?

- Which counties had the most severe sewer overflows (by quantity, time or
both) and when did they occur? Follow-up: Which reported incidents were
highest in terms of overflow that was not collected?

- What is the average quantity of gallons of sewer overflows?

- Can I find the percentage of reported overflows by county? (Imaginary
Example: Wicomico County accounted for 15% of sewer overflows in the
state.)

- What are the most common types of sewer overflows? Follow-up: Could I do
a percentage breakdown of this as well?

- Which Maryland watersheds have been most impacted by sewer overflows?

- (Extra work) How might the frequency or severity of these overflows
compare with Maryland rainfall data?

***Link to the data set:
<https://opendata.maryland.gov/Energy-and-Environment/Reported-Sewer-Overflows/3rgd-zjxx>***

## ***Section #1***

To begin the assignment, I loaded in several libraries that might be relevant, as well as my selected data set, which I cleaned in OpenRefine.

```{r}

# Turn off scientific notation
options(scipen=999)

# Load my libraries
library(tidyverse)
library(janitor)
library(refinr)
library(lubridate)
library(dplyr)
library(readr)

```

```{r}

cleaned_sewer_overflows <- read_csv("data/cleaned_sewer_overflows.csv") |>
  clean_names()

```



