---
title: "grad_assignment"
author: "Pablo Suarez"
date: "2023-11-24"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## INTRODUCTION

Welcome to Pablo Suarez's markdown file for the JOUR772 Graduate Student
Assignment. Below, I am analyzing the Reported Sewer Overflows data set
provided by the Maryland Department of the Environment. This data set
contains information regarding reported sewer overflows across the state
of Maryland from January 1, 2005 through January 31, 2023.

An important note from the data set creators: "Although MDE requires
that all public sewer system owners or operators report overflows to us,
there may be incidents that were not reported. Note that overflow
amounts provided by the person reporting the overflow may be estimated
using best professional judgment or they may be actual readings from
flow measurement devices when available. Penalty information started
during 2013."

Data Profile, Scope and Limitations:

Rows -- 27,479

Columns -- 22

Who created and maintains the data? -- Maryland Department of the
Environment

Dataset Owner -- Andrew Gosden

What state program or law is connected to it? -- Unclear

What does each row represent? -- One overflow event

Columns: overflow type, municipality/facility (it occurred in), NPDES
(National Pollutant Discharge Elimination System identification number
for the facility or system), date discovered, time discovered, duration
(days, hours, minutes), location, zip code, collection-system, quantity
in gallons, net in gallons, cause, watershed, receiving waters, county,
notes, penalty collected, penalty comments, latitude, longitude

What is the chronological and geographic scope of this data? -- Reported
sewer overflows across the state of Maryland from January 1, 2005,
through January 31, 2023.

Newsworthy questions that I'd like to answer with this data set:

-   Which counties reported the most sewer overflows over the dataset's
    observed period of time?

-   Which counties had the most severe sewer overflows (by quantity,
    time or both) and when did they occur? Follow-up: Which reported
    incidents were highest in terms of overflow that was not collected?

-   What is the average quantity of gallons of sewer overflows per event
    versus the estimated net in gallons?

-   Can I find the percentage of reported overflows by county?
    (Imaginary Example: Wicomico County accounted for 15% of sewer
    overflows in the state.)

-   What are the most common types of sewer overflows? Follow-up: Could
    I do a percentage breakdown of this as well?

-   Which Maryland watersheds have been most impacted by sewer
    overflows?

***Link to the data set:
<https://opendata.maryland.gov/Energy-and-Environment/Reported-Sewer-Overflows/3rgd-zjxx>***

## ***Section #1***

To begin the assignment, I loaded in several libraries that I thought
might be relevant, as well as my selected data set, which I cleaned in
OpenRefine. There were a couple of spelling errors that made it through
the OpenRefine process. I corrected them here so that they would not
impact any results.

I also loaded in American Community Survey Maryland county population
data for visualization purposes.

```{r}

# Turn off scientific notation
options(scipen=999)

# Load my libraries
library(tidyverse)
library(janitor)
library(refinr)
library(lubridate)
library(dplyr)
library(readr)
library(stringi)
library(stringr)
library(hash)
library(ggplot2)
library(tigris)
library(tidycensus)
library(sf)
Sys.getenv("CENSUS_API_KEY")
```

```{r}

cleaned_sewer_overflows <- read_csv("data/cleaned_sewer_overflows.csv") |>
  clean_names() 

cleaned_sewer_overflows["county"][cleaned_sewer_overflows["county"] == "Baltimore Ciy"] <- "Baltimore City"
cleaned_sewer_overflows["county"][cleaned_sewer_overflows["county"] == "Batimore City"] <- "Baltimore City"
cleaned_sewer_overflows["county"][cleaned_sewer_overflows["county"] == "Cecil County"] <- "Cecil"
cleaned_sewer_overflows["county"][cleaned_sewer_overflows["county"] == "Harford County"] <- "Harford"
  
glimpse(cleaned_sewer_overflows)

```

```{r}

cleaned_sewer_overflows <- cleaned_sewer_overflows |>
mutate(date_discovered = mdy(date_discovered))

```

```{r}

 md_counties <- get_acs(geography = "county",
    state = "MD",
    year = 2021,
    variables = "B01001_001",
    geometry = TRUE) |>
  clean_names()
                    
head(md_counties)

```

## ***Section #2***

In this section, I attempted to answer the questions listed at the top
of the markdown file.

***Q1: Which counties reported the most sewer overflows over the data
set's observed period of time?***

To answer this question I grouped the data by county, summarized with a
new column called "overflow_count" and arranged the results in
descending order. Immediately, I see that Allegany County has, by far,
the most reported overflows over time with 10,653. The next highest is
Baltimore City with 6,808; followed by Baltimore County with 1,892;
Prince George's County with 1,470 and Montgomery County with 1,146
overflow incidents.

I also included the geoid values for each county and matched them with
the correct county so that I could create a data frame in which I could
join the ACS data by geoid. That required my removal of random values
like Virginia, Loudoun County, VA, and District of Columbia, in order to
be able to merge with the right amount of rows.

```{r}

geoid_mapping <- data.frame(
county = c("Allegany", "Baltimore City", "Baltimore County", "Prince George's", "Montgomery", "Anne Arundel", 
             "Dorchester", "Frederick", "Charles", "Harford", "Howard", "Washington", "Worcester", "St. Mary's", 
             "Talbot", "Wicomico", "Somerset", "Cecil", "Queen Anne's", "Carroll", "Calvert", "Garrett", "Kent", "Caroline"),
geoid = c("24001", "24510", "24005", "24033", "24031", "24003", "24019", "24021", "24017", "24025", "24027", 
            "24043", "24047", "24037", "24041", "24045", "24039", "24015", "24035", "24013", "24009", "24023", "24029", "24011")
)

```

```{r}

overflow_count_geo <- cleaned_sewer_overflows |>
  group_by(county) |>
  summarise(overflow_count = n()) |>
  filter(!(trimws(county) %in% c("Virginia", "Loudoun County, VA", "District of Columbia"))) |>
  filter(county != "NA") |>
  left_join(geoid_mapping, by = "county") |>
  arrange(desc(overflow_count))

```

```{r}

glimpse(overflow_count_geo)

```

As a follow-up to Q1, I took 2021 Maryland population data from the U.S.
Census and joined it to my my overflow_count_geo data frame that I
created. This helped me understand how many residents in each county are
potentially impacted by sewer overflow events. However, I was mostly
interested in getting the geoid and geometry information so that I could
visualize my results on a map. I also wrote this new data frame as a csv
in my data folder so that I could import it into Datawrapper.

```{r}

oc_geopop <- overflow_count_geo |>
  left_join(md_counties, by = "geoid") |>
  select(-c(variable, moe, name))

```

```{r}

write_csv(oc_geopop, "data/oc_geopop.csv")

```

```{r}

glimpse(oc_geopop)

```

***Q2: What is the percentage of reported overflows by county?***

Adding to the last codeblock, this question enables me to show the
percentage breakdown of overflow incidents by county. I did this by
mutating a new column called incident_pct which took the overflow
incident count of a specific county, divided it by the sum of all
overflow incidents and multiplied the results by 100 to get the
percentage.

Unsurprisingly, the results fall in line with what I observed in the
first question. Allegany County had the highest share of incidents with
38.7 percent, followed by Baltimore City with 24.7 percent, then it
drops significantly with Baltimore County at 6.88 percent and Prince
George's County with 5.3 percent.

```{r}

cleaned_sewer_overflows |>
  group_by(county) |>
   summarise(overflow_count = n()) |>
  mutate(incident_pct = overflow_count / sum(overflow_count) * 100) |>
  filter(county != "NA") |>
  arrange(desc(incident_pct))

```

***Q2 Follow-up: What about the distribution of overflows by county and
year?***

I thought this follow-up question would be more straightforward, but I
struggled a lot with getting the results for this question and would
probably need assistance in the future if I were to get to the bottom of
it. I was happy with how the data was formatted when I ran the code
block, but I'm certain that the overflow totals are off because the
total of the Allegany County overflow incidents is nowhere near what I
found in the first question, the same went for other counties. I kept
this code in the project because I felt it was worthwhile to show my
progress and learn from falling short of my goal. It's also entirely
possible that I was overthinking this task.

I tried to complete this question by focusing on finding out how many
overflow incidents each county had in a specific year, a total count of
all overflow incidents across counties for that year and a percentage
breakdown of the share of overflow incidents each county had within each
year. I grouped the data by county and year, with the year value being
extracted from the last four characters of the date column.

Then I calculated the number of distinct dates within each group of
county and year and summarized the results in the column called
'overflow_count.' The n_distinct function in the summarize code helped
me calculate the number of distinct dates within each group of county
and year.

Following that step, I grouped by year and mutated to create two new
columns. The first column calculated the total number of incidents
within each year by adding the overflow count. The second column
calculated each county's percentage of overflow incidents by year in
relation to all overflow incidents across counties for that year.
Finally, the data is arranged by county and year.

I also think that it would be worthwhile and potentially newsworthy to
go even further and compute the percentage of a county's overflow
incidents from a certain year in relation to its total number of
incidents over all recorded years in the data set. It would help me
understand which counties had particularly active years in terms of
overflow incidents in relation to other counties. It would also help me
complete a historical comparison of the overflow incidents from a
specific year within a county and its historical total.

Here is the code that I tried to use:

```{r}

cleaned_sewer_overflows |>
  group_by(county, year = substr(date_discovered, nchar(date_discovered) - 3, nchar(date_discovered))) |>
  summarise(overflow_count = n_distinct(date_discovered)) |>
  group_by(year) |>
  mutate(total_count_year = sum(overflow_count),
         incident_pct_that_year = overflow_count / total_count_year * 100) |>
  arrange(county, year)

```

***Q3: Which counties had the most severe sewer overflows (by quantity,
time or both) and when did they occur?***

To find the answer to this question, I grouped the data by county and
date discovered. I then only kept the columns I was interested in
seeing: date_discovered, duration_days, hours, minutes,
quantity_in_gallons_estimated, cause and county. Then I arranged the
data in descending order based on duration and
quantity_in_gallons_estimated.

The longest sewer overflow event, according to the data, occurred in
Dorchester County on Sept. 25, 2006. This event in particular lasted 15
years (5,475 days) and had an estimated 702,000 gallons of overflow
water. The sewer overflow event with the highest estimated quantity in
gallons occurred in Allegany County in 2014 with 106,102,592 gallons.
This event did not have duration data that we could measure against the
other event.

```{r}

cleaned_sewer_overflows |>
  group_by(county, date_discovered) |> 
  select(-c(overflow_type, municipality_facility, npdes_number, location, zip_code, collection_system, net_in_gallons_estimated, watershed, receiving_waters, notes, penalty_collected, latitude, longitude, penalty_comments, time_discovered)) |>
  arrange(desc(duration_days), desc(duration_hours), desc(duration_minutes), desc(quantity_in_gallons_estimated))
  #arrange(desc(quantity_in_gallons_estimated))

```

As a follow-up to that question, I followed the same process to identify
the counties with overflow events highest in net estimated gallons. Once
again, the answer was the 2014 Allegany County incident.

```{r}

cleaned_sewer_overflows |>
  group_by(county, date_discovered) |> 
  select(-c(overflow_type, municipality_facility, npdes_number, location, zip_code, collection_system, watershed, receiving_waters, notes, penalty_collected, latitude, longitude, penalty_comments, time_discovered, quantity_in_gallons_estimated)) |>
arrange(desc(net_in_gallons_estimated))

```

***Q4: What is the average quantity of gallons of sewer overflows per
event versus the estimated net in gallons?***

There wasn't much code needed to calculate this value. I added the
quantity_in_gallons_estimated column while removing any NA values and
then divided by the total number of incidents. I then did same for the
net_in_gallons_estimated column to see if there was a significant
difference from the first column. I commented the answers within each
code block below.

```{r}

sum(cleaned_sewer_overflows$quantity_in_gallons_estimated, na.rm = TRUE) / 27479 
#Answer is 255380.5 gallons per event

```

```{r}

sum(cleaned_sewer_overflows$net_in_gallons_estimated, na.rm = TRUE) / 27479 
#Answer is 254185.4 gallons per event

```

***Q5: Which Maryland watersheds have been most impacted by sewer
overflows?***

A group_by, summarize and filtering of "NA" results led me to acquiring
the count of incidents occurring in specific watersheds. The highest
amount was "Unknown" watersheds, followed by George's Creek as the
highest named watershed.

```{r}

cleaned_sewer_overflows |>
  group_by(receiving_waters) |>
  summarise(rw_incidents = n()) |> 
  filter(receiving_waters != "NA|N/A") |>
  arrange(desc(rw_incidents))

```

***Q6: What are the most common types of sewer overflows?***

First, I made a new data frame that isolates blockage causes and a total
count. I wasn't satisfied with this raw count because I was more
interested in finding out the number of times they were either the sole
and/or contributing factor in a sewer blockage.

```{r}

blockage_causes <- cleaned_sewer_overflows |>
  group_by(cause) |>
  summarise(cause_count = n()) |>
  filter(cause != "NA|N/A") |>
  mutate(cause = str_to_lower(cause)) |>
  arrange(desc(cause_count))

```

One method to accomplish this would be to filter the cause column in
this data frame and string detect each major cause. However, this method
would not accurately track causes nested within rows that have multiple
causes contributing to a blockage.

```{r}
#Don't need to run this code block
blockage_causes |> 
  filter(str_detect(cause, "Precipitation")) |> 
    summarise(cause = "Precipitation", total = sum(cause_count))

```

Instead, I opted to create a 'for' loop. I first created a list of the
most common factors contributing to sewer blockages and overflows. The
next line of code "h = hash()" creates an empty hash in which we store
key value pairs where the key is the 'cause' of the sewer overflow and
the value is the count.

From there, I created the loop checking the data frame for each item in
the list. Essentially, this loop is string detecting for each of these
causes, then summarizing the count in which they appear. Finally, I
wrote code to turn the result of the for loop into a separate data frame
called 'cause_counts' that I could use to conduct further analysis.

```{r}

cause_list <- list("rags", "grease", "roots", "rain", "trash", "debris", "sand", "gravel", "rocks", "break", "malfunction", "excess", "fail", "vandalism", "defective", "error", "precipitation")

h = hash()

for (x in cause_list) {
 df =  blockage_causes |> 
    filter(str_detect(cause, x)) |> 
    summarise(cause = x, total = sum(cause_count))

  h[[x]] <- df$total

}

```

```{r}

hash_list <- list(causes = keys(h), counts = values(h))
cause_counts <- as.data.frame(hash_list)

```

With this last code block below, I can see that precipitation, rain,
rags, grease and roots are the top five culprits. Precipitation and rain
could mean the same thing, however, it's tough to know exactly how each
were meant to be interpreted in the original data frame so I left them
apart. I then took the total number of incidents (rows) from the
original data frame and divided each count by that number to find the
frequency in which these terms appeared in the cause column. I then
wrote that new data frame as a CSV file so that I could import it into
Datawrapper for visualization.

```{r}

cause_counts |>
  mutate(pct_cause = counts / (27479) * 100) |>
  arrange(desc(pct_cause))

```

```{r}

write_csv(cause_counts, "data/cause_counts.csv")

```

## ***Section #3***

This final section contains potential story ideas and my two data
visualization examples.

Chart: "Top Contributing Causes of Maryland Sewer Overflow Incidents"
<https://datawrapper.dwcdn.net/2E9nb/1/>

```{r}

cause_counts |>
  ggplot() +
  geom_bar(aes(x = reorder(causes, -counts), weight = counts)) +
  labs(
    title = "Top Contributing Causes of Maryland Sewer Overflow Incidents",
    x = "Overflow Cause",
    y = "Number of Mentions",
    caption = "Source: Maryland Department of the Environment") +
  theme(
    axis.text.x = element_text(angle = 45,  hjust=1))

```

Map: "Counties Reporting Sewer Overflows Between 2005-2023"
<https://datawrapper.dwcdn.net/4d31g/1/>

```{r}

ggplot() +
  geom_sf(data = oc_geopop, aes(fill = overflow_count, geometry = geometry)) +
  scale_fill_viridis_c(option = "magma") +  
  theme_minimal() +
  labs(
    title = "Counties Reporting Sewer Overflows Between 2005-2023",
    caption = "Source: Maryland Department of the Environment") 

```

Story Ideas and Questions Worth Exploring:

-   #1: What's going on in Allegany County and why does it have so many
    more sewer overflow incidents than the rest of the state? Over the
    time period that the data set covers, Allegany County saw
    significantly more overflow incidents compared to other counties.
    With over 10,000 reported sewer overflows over roughly 18 years, it
    had nearly 4,000 more incidents than the next closest county,
    Baltimore City. This story could be a deep dive into the county by
    looking into its events to determine their severity and whether the
    frequency of its incidents have increased, decreased or remained
    consistent over time. Also, what are specific problem areas within
    the county? We would map those results, as well as identify whether
    these incidents are typically caused by the same factors.

-   #2: Precipitation and rainfall are the top contributors to sewer
    overflow incidents in Maryland. How might Maryland counties fare in
    the future? For this story, I'd want to see the frequency or
    severity of overflow incidents compared to Maryland rainfall data
    and events over the years. Does the rainfall data warrant the amount
    of overflow incidents in certain Maryland counties? And for 2024,
    what are the projections among meteorologists regarding expected
    rainfall and which counties should be prepared to bolster their
    sewage systems? In an earlier code block, I showed the percentage
    breakdown of contributing overflow causes in relation to the whole
    original data frame. However, one avenue potentially worth exploring
    is going even further and calculating that percentage on a
    county-by-county basis (i.e., what were the most prevalent overflow
    causes by county and could we pin down specific years during which
    these causes were more prevalent?).

-   #3: What's the story with the 2006 overflow incident in Dorchester
    County? This could have been an error on either the part of the
    person who reported the incident or someone who imported the data
    into the data set. I couldn't find a story online about this
    overflow that lasted basically 15 years. Even though it wasn't the
    worst event in terms of the overflow quantity in gallons, it still
    seems worthwhile to find out what happened there. If the data is
    correct, it would be nice to follow up with a county official to
    figure out what work they completed to stop the overflow and gain a
    better understanding of how the county has updated its sewer
    infrastructure.

-   #4: I wrote about my Q2 follow-up question shortcomings in an
    earlier code block and I think it's potentially a worthwhile story
    to explore. Please refer to that section to read my reasoning.

-   #5: My classmate, Adriana Navarro, recommended that I speak with UMD
    professor Marccus Hendricks about my project. Hendricks is an
    associate professor of urban studies and (environmental) planning
    and is the director of the Stormwater Infrastructure Resilience and
    Justice (SIRJ) Lab. It would be interesting to get his perspective
    about my results from identifying the most common causes of sewer
    overflows and how they compare to any research that he's conducted
    into the same issue. Here is a link to Hendrick's faculty bio:
    <https://arch.umd.edu/people/marccus-hendricks>
